---
title: "Summary Stats"
output:
  word_document: default
  html_notebook: default
  pdf_document: default
---


```{r include=FALSE}
library(tidyverse)
library(gt)
library(here)
```


```{r include=FALSE}
ortho_data <- read_csv(here("Data", "2025_04_10_Ortho_Data_with_groups.csv"))
```


### 1. Number of Children in Each Group

The table below provides information on the number of participants in each group.

```{r echo=FALSE}
# Number of children in each group

children_by_group <- ortho_data %>%
  distinct(participant_id, Group) %>%
  count(Group, name = "n_children") %>%
  gt() %>%
  cols_label(
    Group = "Group",
    n_children = "Number of Participants (n)") %>%
 tab_header(
    title = "Table 1: Number of participants in each group"
  )

children_by_group
```

### 2. Trials Where Participants Fixated on All Four Interest Areas (IAs)

The first inclusion criteria indicates that the participant must have looked at all four interest areas. Table 2 provides information on the number of trials that each participant (per group) fixated on all four IAs.

```{r echo=FALSE}

# Step 1: Create trial level data based on Trial index

trial_level_data <- ortho_data %>%
  mutate(Ortho_ET_IA_inclusion = as.numeric(Ortho_ET_IA_inclusion)) %>%  
  group_by(Group, participant_id, OrthoET_trial_number) %>%
  summarise(
    fixated_all_four = sum(Ortho_ET_IA_inclusion) == 4,  
    .groups = "drop"
  )

# Step 2: Count how many trials were fixated vs not, per participant
fixation_counts <- trial_level_data %>%
  group_by(Group, participant_id) %>%
  summarise(
    `Number of Trials Participants Fixated on All IAs` = sum(fixated_all_four),
    `Number of Trials Participants DID NOT Fixate on All IAs` = sum(!fixated_all_four),
    .groups = "drop"
  ) %>%
  rename(ID = participant_id)

fixation_counts %>%
  gt()

```

### 3. Accuracy of Trials per Participant

The second inclusion criteria indicates that the participant must have been accurate in at least 12 trials (that is, they selected either the high- or low-probability orthographic nonword over the illegal or unpronounceable foil on at least 12 trials). Table 2 provides information on the number of trials that each participant was accurate on.

```{r echo=FALSE}
# Step 1: Ensure accuracy column is numeric (in case it's "1"/"0" as string)
ortho_data <- ortho_data %>%
  mutate(Ortho_ET_accuracy = as.numeric(Ortho_ET_accuracy))

# Step 2: Collapse to one row per trial
# Here we assume that if any IA in a trial is marked inaccurate (0), the whole trial is inaccurate
# Or change logic as needed depending on your task
trial_accuracy <- ortho_data %>%
  group_by(Group, participant_id, OrthoET_trial_number) %>%
  summarise(
    trial_accurate = all(Ortho_ET_accuracy == 1),  
    .groups = "drop"
  )

# Step 3: Count accurate and inaccurate trials per participant
accuracy_summary <- trial_accuracy %>%
  group_by(Group, participant_id) %>%
  summarise(
    `Accurate` = sum(trial_accurate),
    `Inaccurate` = sum(!trial_accurate),
    .groups = "drop"
  ) %>%
  rename(ID = participant_id)

# Step 4: View summary table
accuracy_summary %>%
  gt()

```


### 4. Number of "Good" Trials

A “good” trial was defined as one in which the participant (a) fixated on all four interest areas (IAs), and (b) made an accurate response on at least 12 trials — that is, they selected either the high- or low-probability orthographic nonword over the illegal or unpronounceable foil on at least 12 trials

The table below shows the number and percentage of participants excluded due to either having fewer than 12 good trials or not fixated on all four IAs, across the three groups:


```{r echo=FALSE}
library(gt)
# Step 1: Collapse to one row per participant per trial
trial_level <- ortho_data %>%
  group_by(participant_id, Group, OrthoET_trial_number) %>%
  summarise(
    trial_met_both = any(Ortho_ET_met_both_inclusion == 1),
    .groups = "drop"
  )
# Start from your cleaned trial-level summary
participant_table <- trial_level %>%
  group_by(participant_id, Group) %>%
  summarise(
    n_good_trials = sum(trial_met_both),
    met_both_inclusion = n_good_trials >= 12,
    .groups = "drop"
  ) %>%
  mutate(
    inclusion_status = ifelse(met_both_inclusion, "Passed", "Failed")
  )

# Display in a gt table
participant_table %>%
  select(Group, participant_id, n_good_trials, inclusion_status) %>%
  gt() %>%
  cols_label(
    Group = "Group",
    participant_id = "Participant ID",
    n_good_trials = "Number of Good Trials",
    inclusion_status = "Inclusion Status"
  ) %>%
  tab_header(
    title = "Participant Inclusion Based on Accuracy and Fixation Criteria"
  )

```

Among participants excluded, it was examined whether their exclusion was primarily due to failure to meet the fixation criterion rather than poor task performance. Specifically, participants who were accurate on at least 12 trials (i.e., selected either a high- or low-orthographic nonword), but were still excluded because they did not meet the requirement of fixating on all four interest areas (IAs) were identified.

The table below summarizes how many behaviorally accurate participants were excluded from each group:

```{r echo=FALSE}
trial_level_detail <- ortho_data %>%
  group_by(participant_id, Group, OrthoET_trial_number) %>%
  summarise(
    trial_accurate = any(Ortho_ET_accuracy == 1),
    fixated_all_IAs = sum(Ortho_ET_IA_inclusion == 1) == 4,
    .groups = "drop"
  )
participant_reasons <- trial_level_detail %>%
  group_by(participant_id, Group) %>%
  summarise(
    n_accurate_trials = sum(trial_accurate),
    n_fixated_trials = sum(fixated_all_IAs),
    .groups = "drop"
  ) %>%
  mutate(
    accuracy_inclusion = n_accurate_trials >= 12,
    fixation_inclusion = n_fixated_trials >= 12
  )
participant_diagnostics <- participant_table %>%
  left_join(participant_reasons, by = c("participant_id", "Group"))
participant_diagnostics %>%
  filter(inclusion_status == "Failed") %>%
  select(
    Group, participant_id,
    n_good_trials,
    n_accurate_trials, accuracy_inclusion,
    n_fixated_trials, fixation_inclusion
  ) %>%
  gt() %>%
  cols_label(
    participant_id = "Participant ID",
    n_good_trials = "Good Trials (Both Criteria)",
    n_accurate_trials = "Accurate Trials",
    accuracy_inclusion = "Passed Accuracy?",
    n_fixated_trials = "Fixated Trials",
    fixation_inclusion = "Passed Fixation?"
  ) %>%
  tab_header(title = "Why Did These Participants Fail Inclusion?")


```
Since participants c171082 and c194017 did not pass the inclusion criteria, they were removed for further analysis.

## 



































































