---
title: "Orthographic Processing in School-age Children - Descriptive Statistics Eye Tracking Data"
format: typst
editor: visual
---

### Brief Introduction
\
Orthographic awareness (OA) — the ability to recognize permissible letter patterns in a given writing system—is an important component of reading development. In alphabetic languages such as English, OA supports both decoding and word recognition by enabling readers to distinguish between legal and illegal letter sequences. The current study investigates orthographic processing in school-age children (grades 3–6) across three groups: children with dyslexia (n = 28), children with compensated dyslexia (n = 17), and children with typical development (TD, n = 23). By examining how these groups differ in their orthographic decision-making, the study aims to better understand the mechanisms underlying persistent and resolved reading difficulties.

To assess OA, participants completed a visual world eye-tracking task in which they were asked: **"Which word looks most like a real English word?"** Each trial presented four non-word options that varied in orthographic probability and legality:

- A high-probability (high-ortho) item resembling real English words,

- A low-probability (low-ortho) item with less typical letter sequences,

- An illegal item that violates English orthographic rules,

- An unpronounceable item with no plausible phonological form.

Participants completed 18 trials, with option locations and trial order randomized. The task was administered using E-Prime with eye-tracking data collected concurrently.

This report investigates group differences across multiple dimensions of task performance, including:

1. Response time when selecting high-ortho and low-ortho items,

2. Accuracy, based on criteria where high-ortho or both high/low-ortho selections are considered correct,

3. Eye movement metrics, such as fixation count and  total dwell time

4. Orthographic interference ratios, including the relative attention (via dwell time) given to legal vs. illegal options.


### Inclusion Criteria
\
Participants were included in the analyses if they met both of the following eye-tracking experiment criteria:

- Accuracy: Correctly identified at least 67% of trials (≥12/18), using the high/low orthographic items as correct choices.

- Visual Engagement: Had at least 12 trials in which they visually inspected all four options (i.e., no fixation count value was zero across the four interest areas).

This inclusion process was implemented to ensure that participants understood the task instructions and attended to all the presented stimuli before making their selection. By applying these filters, the aim was to increase the reliability of the response time and eye-tracking measures used in subsequent analyses.


```{r}
#| include: false
## Loading Libraries
library(here)
library(tidyverse)
library(gt)
```


```{r}
#| include: false
## Loading File

ortho_data <- read_csv(here("Data", "2025_04_10_Ortho_Data_final.csv"))
```

Based on this inclusion criteria, the TD group had 22 participants, compensated group had 16 participants and the dyslexia group had 28 participants.

\

### Descriptive Statistics
\

#### 1. Response Time (RT) when High_ortho is selected
\

```{r}
#| echo: false

# Step 1: Filter for High_Ortho responses
high_ortho_rt <- ortho_data %>%
  filter(OrthoET_response_choice == "High_Ortho")

# Step 2: Deduplicate per participant × trial
high_ortho_rt_unique <- high_ortho_rt %>%
  distinct(participant_id, OrthoET_trial_number, .keep_all = TRUE)

## Reordering groups

high_ortho_rt_unique <- high_ortho_rt_unique %>%
  mutate(Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated")))

# Step 3: Summary statistics with response time in seconds
high_ortho_summary <- high_ortho_rt_unique %>%
  group_by(Group) %>%
  summarise(
    Mean = mean(OrthoET_response_time, na.rm = TRUE) / 1000,
    SD = sd(OrthoET_response_time, na.rm = TRUE) / 1000,
    Min = min(OrthoET_response_time, na.rm = TRUE) / 1000,
    Q1 = quantile(OrthoET_response_time, 0.25, na.rm = TRUE) / 1000,
    Median = median(OrthoET_response_time, na.rm = TRUE) / 1000,
    Q3 = quantile(OrthoET_response_time, 0.75, na.rm = TRUE) / 1000,
    Max = max(OrthoET_response_time, na.rm = TRUE) / 1000,
    .groups = "drop"
  )

# Number of total trials per group
high_ortho_rt_unique %>%
  count(Group, name = "Total_Trials")%>%
  gt()

# Number of trials per participant
high_trials_per_participant <- high_ortho_rt_unique %>%
  count(participant_id, Group, name = "Trials_Selected_High_Ortho")

# summary (min/max/mean) number of High_Ortho selections per participant by group
high_trials_per_participant %>%
  group_by(Group) %>%
  summarise(
    Min = min(Trials_Selected_High_Ortho),
    Max = max(Trials_Selected_High_Ortho),
    Mean = mean(Trials_Selected_High_Ortho),
    SD = sd(Trials_Selected_High_Ortho),
    .groups = "drop"
  ) %>%
  gt()

```

For the response time analysis involving trials where participants selected the High_Ortho option, a total of 302 trials were included from the TD group, 350 from the dyslexia group, and 211 from the compensated group. On average, participants contributed between 12 and 14 trials each (TD: M = 13.73, SD = 1.86; Dyslexia: M = 12.50, SD = 2.32; Compensated: M = 13.19, SD = 2.83), with the number of High_Ortho selections ranging from 7 to 17 trials.

Descriptive statistics for response time (in seconds) on trials where participants selected the high-orthographic probability item are presented in Table 1. 

```{r}
#| echo: false

# table
high_ortho_summary %>%
  gt() %>%
  tab_header(
    title = "Table 1: Response Time Summary (in seconds) for High_Ortho Selections",
    subtitle = "By Reading Group"
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 2
  )%>%
  tab_options(
    heading.title.font.size = 14,     # Default is 24
    heading.subtitle.font.size = 12   # Default is 16
  )


```

The TD group had a mean response time of 5.86 seconds (SD = 3.14), the dyslexia group averaged 6.15 seconds (SD = 2.95), and the compensated dyslexia group averaged 5.23 seconds (SD = 2.59). Median response times were 5.25, 5.51, and 4.74 seconds for the TD, dyslexia, and compensated groups, respectively. The interquartile ranges were similar across groups: 3.91–6.89 seconds for TD, 4.52–6.97 seconds for dyslexia, and 3.65–6.13 seconds for the compensated group. Although all groups included some long response times (e.g., maximums exceeding 19 seconds), overall performance patterns suggest comparable timing across groups when selecting high-orthographic items.

```{r}
#| echo: false
high_ortho_rt_unique <- high_ortho_rt_unique %>%
  mutate(
    RT_sec = OrthoET_response_time / 1000,
    Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated"))
  )

ggplot(high_ortho_rt_unique, aes(x = Group, y = RT_sec)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Response Times (High_Ortho Selections)",
    y = "Response Time (s)",
    x = "Group"
  ) +
  theme_minimal()

high_ortho_rt_unique %>%
  filter(RT_sec > 15) %>%
  arrange(desc(RT_sec)) %>%
  select(participant_id, Group, OrthoET_trial_number, RT_sec)

```

Visual inspection of the boxplot revealed several response times above 15 seconds in each group, suggesting the presence of potential outliers. These long RTs were retained in the analysis as they may reflect natural variation in decision-making time rather than noise. However, they may warrant further examination in subsequent analyses.

\

#### 2. Response Time (RT) when Low_ortho is selected

\
For Low_Ortho selections, fewer trials contributed to the response time analysis, with 91 trials from the TD group, 145 from the dyslexia group, and 77 from the compensated group. Participants selected the Low_Ortho option on average 4 to 5 times (TD: M = 4.14, SD = 1.83; Dyslexia: M = 5.18, SD = 1.83; Compensated: M = 4.81, SD = 2.83), with a minimum of 1 and a maximum of 10 selections per participant.

```{r}
#| echo: false
# Step 1: Filter for Low_Ortho responses
low_ortho_rt <- ortho_data %>%
  filter(OrthoET_response_choice == "Low_Ortho")

# Step 2: Deduplicate per participant × trial
low_ortho_rt_unique <- low_ortho_rt %>%
  distinct(participant_id, OrthoET_trial_number, .keep_all = TRUE)

# Step 3: Reorder groups and convert RT to seconds
low_ortho_rt_unique <- low_ortho_rt_unique %>%
  mutate(
    Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated")),
    RT_sec = OrthoET_response_time / 1000
  )

# Step 4: Summary statistics
low_ortho_summary <- low_ortho_rt_unique %>%
  group_by(Group) %>%
  summarise(
    Mean = mean(RT_sec, na.rm = TRUE),
    SD = sd(RT_sec, na.rm = TRUE),
    Min = min(RT_sec, na.rm = TRUE),
    Q1 = quantile(RT_sec, 0.25, na.rm = TRUE),
    Median = median(RT_sec, na.rm = TRUE),
    Q3 = quantile(RT_sec, 0.75, na.rm = TRUE),
    Max = max(RT_sec, na.rm = TRUE),
    .groups = "drop"
  )

# Total trials per group
low_ortho_rt_unique %>%
  count(Group, name = "Total_Trials") %>%
  gt()

# Trials per participant
low_trials_per_participant <- low_ortho_rt_unique %>%
  count(participant_id, Group, name = "Trials_Selected_Low_Ortho") 

# Summary by group
low_trials_per_participant %>%
  group_by(Group) %>%
  summarise(
    Min = min(Trials_Selected_Low_Ortho),
    Max = max(Trials_Selected_Low_Ortho),
    Mean = mean(Trials_Selected_Low_Ortho),
    SD = sd(Trials_Selected_Low_Ortho),
    .groups = "drop"
  )%>%
  gt()



```

Descriptive statistics for response time (in seconds) on trials where participants selected the low-orthographic probability item are presented in Table 2.

```{r}
#| echo: false
# Step 5: Print with gt
low_ortho_summary %>%
  gt() %>%
  tab_header(
    title = "Table 2: Response Time Summary (in seconds) for Low_Ortho Selections",
    subtitle = "By Reading Group"
  ) %>%
  fmt_number(columns = everything(), decimals = 2)%>%
  tab_options(
    heading.title.font.size = 14,     # Default is 24
    heading.subtitle.font.size = 12   # Default is 16
  )
```


The typical development (TD) group had a mean response time of 7.32 seconds (SD = 6.29), while the dyslexia and compensated dyslexia groups had means of 7.19 seconds (SD = 3.87) and 5.70 seconds (SD = 2.65), respectively. The TD group also exhibited the widest range of response times, with a maximum of 50.78 seconds.

```{r}
#| echo: false
ggplot(low_ortho_rt_unique, aes(x = Group, y = RT_sec)) +
  geom_boxplot() +
  labs(title = "Boxplot of Response Times (Low_Ortho Selections)",
       y = "Response Time (s)", x = "Group") +
  theme_minimal()

low_ortho_rt_unique %>%
  filter(Group == "TD" & RT_sec > 20) %>%
  arrange(desc(RT_sec)) %>%
  select(participant_id, OrthoET_trial_number, RT_sec)
```

Visual inspection of boxplots indicated the presence of a few unusually long trials across all groups, particularly in the TD group, suggesting potential outliers. These values were retained in the analysis, but they may warrant further consideration in follow-up analyses.
\

#### 3. Accuracy scores by group when High_ortho was selected

\
Descriptive statistics for accuracy based on the selection of high-ortho probability items are presented in Table 3.

```{r}
#| echo: false
## accuracy across groups
accuracy_high_summary <- ortho_data %>%
  distinct(participant_id, OrthoET_trial_number, .keep_all = TRUE) %>%  # one row per trial
  mutate(Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated"))) %>%
  group_by(Group) %>%
  summarise(
    Mean_Accuracy = mean(OrthoET_response_choice_accuracy_high, na.rm = TRUE),
    SD = sd(OrthoET_response_choice_accuracy_high, na.rm = TRUE),
    .groups = "drop"
  )
## Table
accuracy_high_summary %>%
  gt() %>%
  tab_header(
    title = "Table 3: Accuracy for High_Ortho Selections Only",
    subtitle = "Proportion of Trials Where High_Ortho Was Selected (Out of 18)"
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  ) %>%
  fmt_number(columns = everything(), decimals = 2)

```
Accuracy was calculated as the proportion of trials (out of 18) on which each participant selected the High_Ortho option. The typical development (TD) group demonstrated the highest mean accuracy (M = 0.76, SD = 0.43), followed by the compensated group (M = 0.73, SD = 0.44) and the dyslexia group (M = 0.69, SD = 0.46). 
\

#### 4. Accuracy Scores by Group when High and Low Ortho were selected

\
Descriptive statistics for accuracy using a broader criterion—counting both high- and low-orthographic probability selections as correct—are presented in Table 4.

```{r}
#| echo: false
## Accuracy high and low ortho
accuracy_highlow_summary <- ortho_data %>%
  distinct(participant_id, OrthoET_trial_number, .keep_all = TRUE) %>%
  mutate(Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated"))) %>%
  group_by(Group) %>%
  summarise(
    Mean_Accuracy = mean(Ortho_ET_accuracy, na.rm = TRUE),
    SD = sd(OrthoET_response_choice_accuracy_highlow, na.rm = TRUE),
    .groups = "drop"
  )

## table

accuracy_highlow_summary %>%
  gt() %>%
  tab_header(
    title = "Table 4: Accuracy for High or Low_Ortho Selections",
    subtitle = "Proportion of Trials Where Either High or Low_Ortho Was Selected (Out of 18)"
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  ) %>%
  fmt_number(columns = everything(), decimals = 2)

```

All groups performed near ceiling under this scoring approach. The typical development (TD) group achieved a mean accuracy of 0.99 (SD = 0.09), the dyslexia group had a mean of 0.98 (SD = 0.13), and the compensated group performed at 1.00 (SD = 0.00). This pattern suggests that across reading profiles, children were highly likely to select one of the two orthographically legal options on each trial.

\

#### 5. Proportion of fixations on each IA type

\

Table 5 presents the mean proportion of fixations on each response option, calculated as the percentage of total fixations per trial.

```{r}
#| echo: false
# Step 1: Compute total fixations per trial (participant + trial)
ortho_percent_fix <- ortho_data %>%
  filter(!is.na(OrthoET_IA_fixation_count)) %>%
  group_by(participant_id, OrthoET_trial_number) %>%
  mutate(
    total_fixations_trial = sum(OrthoET_IA_fixation_count, na.rm = TRUE),
    fixation_percent = OrthoET_IA_fixation_count / total_fixations_trial
  ) %>%
  ungroup()
## Relabeling 
ortho_percent_fix <- ortho_percent_fix %>%
  mutate(
    OrthoET_IA_label = recode(
      OrthoET_IA_label,
      "High_Ortho_IA" = "High Ortho",
      "Low_Ortho_IA" = "Low Ortho",
      "Illegal_IA" = "Illegal",
      "Unpron_IA" = "Unpronounceable"
    )
  )

## Summarize percent fixations by group and IA type

fixation_percent_summary <- ortho_percent_fix %>%
  mutate(
    Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated")),
    OrthoET_IA_label = factor(OrthoET_IA_label, levels = c("High Ortho", "Low Ortho", "Illegal", "Unpronounceable"))
  ) %>%
  group_by(Group, OrthoET_IA_label) %>%
  summarise(
    Mean_Fixation_Percent = mean(fixation_percent, na.rm = TRUE) * 100,
    SD = sd(fixation_percent, na.rm = TRUE) * 100,
    .groups = "drop"
  )

```


```{r}
#| echo: false
library(dplyr)
library(tidyr)
library(gt)
library(glue)

# Combine mean and SD into one formatted cell
fixation_percent_formatted <- fixation_percent_summary %>%
  mutate(
    value = glue("{round(Mean_Fixation_Percent, 2)}\n({round(SD, 2)})")
  ) %>%
  select(Group, OrthoET_IA_label, value)

# Pivot wider
fixation_percent_table <- fixation_percent_formatted %>%
  pivot_wider(
    names_from = OrthoET_IA_label,
    values_from = value
  )

# Display using gt
fixation_percent_table %>%
  gt() %>%
  tab_header(
    title = "Table 5: Proportion of Fixations by Option Type",
    subtitle = "Mean (SD) % of Fixations per Group"
  ) %>%
  tab_style(
    style = cell_text(size = "smaller"),
    locations = cells_body()
  )%>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  ) 

```

Across all groups, participants fixated most frequently on the high-orthographic option (TD: 35.27%, Dyslexia: 33.14%, Compensated: 33.92%), followed by the low-orthographic option. Illegal and unpronounceable items received fewer fixations overall, with proportions typically under 20%. Standard deviations were relatively consistent across groups

\

#### 6. Total Dwell Time on each option

\

Table 6 presents the average dwell time (in seconds) for each response option type.

```{r}
#| echo: false
# Prepare data
ortho_dwell_summary <- ortho_data %>%
  filter(!is.na(OrthoET_IA_dwell_time)) %>%
  mutate(
    Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated")),
    OrthoET_IA_label = recode(
      OrthoET_IA_label,
      "High_Ortho_IA" = "High Ortho",
      "Low_Ortho_IA" = "Low Ortho",
      "Illegal_IA" = "Illegal",
      "Unpron_IA" = "Unpronounceable"
    ),
    OrthoET_IA_label = factor(
      OrthoET_IA_label,
      levels = c("High Ortho", "Low Ortho", "Illegal", "Unpronounceable")
    ),
    dwell_time_sec = OrthoET_IA_dwell_time / 1000
  ) %>%
  group_by(Group, OrthoET_IA_label) %>%
  summarise(
    Mean = mean(dwell_time_sec, na.rm = TRUE),
    SD = sd(dwell_time_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    value = glue("{round(Mean, 2)} ({round(SD, 2)})")
  )

```

```{r}
#| echo: false

# Reshape to wide format
dwell_time_table <- ortho_dwell_summary %>%
  select(Group, OrthoET_IA_label, value) %>%
  pivot_wider(
    names_from = OrthoET_IA_label,
    values_from = value
  )

# Display
dwell_time_table %>%
  gt() %>%
  tab_header(
    title = "Table 6: Dwell Time on Each Option Type",
    subtitle = "Mean (SD) Dwell Time in Seconds per Group"
  ) %>%
  tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  )

```

All groups spent the most time viewing the high-orthographic item, but the dyslexia group showed notably longer dwell times for low-orthographic, illegal, and unpronounceable items compared to the other groups. For example, mean dwell time on low-orthographic items was 1.98 seconds for the dyslexia group, versus 1.63 and 1.60 seconds for the TD and compensated groups, respectively. A similar trend was observed for illegal and unpronounceable options. 

#### 7. Ratio of Total Dwell Time on Legal vs. Illegal Options



```{r}
## Classify IA as legal and illegal
dwell_ratio_data <- ortho_data %>%
  filter(!is.na(OrthoET_IA_dwell_time)) %>%
  mutate(
    dwell_time_sec = OrthoET_IA_dwell_time / 1000,
    Legal_Status = case_when(
      OrthoET_IA_label %in% c("High_Ortho_IA", "Low_Ortho_IA") ~ "Legal",
      OrthoET_IA_label %in% c("Illegal_IA", "Unpron_IA") ~ "Illegal",
      TRUE ~ NA_character_
    )
  )

## Computing Ratio

# Sum dwell time by participant × trial × Legal status
dwell_ratio_trial <- dwell_ratio_data %>%
  group_by(participant_id, OrthoET_trial_number, Group, Legal_Status) %>%
  summarise(
    total_dwell = sum(dwell_time_sec, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Legal_Status,
    values_from = total_dwell
  ) %>%
  mutate(
    dwell_ratio_legal_vs_illegal = Legal / Illegal
  )

## Summarize ratio by group

dwell_ratio_summary <- dwell_ratio_trial %>%
  mutate(Group = factor(Group, levels = c("TD", "Dyslexia", "Compensated"))) %>%
  group_by(Group) %>%
  summarise(
    Mean_Ratio = mean(dwell_ratio_legal_vs_illegal, na.rm = TRUE),
    SD = sd(dwell_ratio_legal_vs_illegal, na.rm = TRUE),
    .groups = "drop"
  )

## table

dwell_ratio_summary %>%
  gt() %>%
  tab_header(
    title = "Ratio of Dwell Time on Legal vs. Illegal Options",
    subtitle = "Mean (SD) Dwell Time Ratio by Group"
  ) %>%
  fmt_number(columns = everything(), decimals = 2)


```







